E-graphs With Bindings

### 2025-01-18 LICS_2025/e_graphs_with_bindings/E-graphs-with-bindings.tex
e-graphs

E-graphs are a data structure which can efficiently represent many equivalent terms of an algebraic theory.  They generalise the typical DAG representation of terms to include equivalence classes.
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Figure~\ref{fig:e-graph-example}.
This example is a `Hello world!' of e-graphs, presented first in~\cite{EggPaper}.

@@

Each column represents a term (or term rewrite rule), the conventional e-graph representation, and the equivalent string diagram representation, an e-hypergraph as developed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}. 
The initial term, $(a*2)/2$ (which corresponds to subfigure (a)), is already represented efficiently in the e-graph by sharing the node 2.
The string diagram version is slightly different than conventional e-graphs by representing the sharing of the node 2 explicitly, using a sharing (contraction) node
\scalebox{0.3}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

Nodes of an e-graph are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
Initially each equivalence class has a single node (subgraph).
The first rewrite rule, replacing multiplication by the more efficient \emph{shift-left} operator creates the first non-unitary (containing more than one node or subgraph) equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) operator nodes, with a new sharing, that of node $a$.
This second e-graph illustrates the other difference, besides sharing, between conventional e-graphs and e-hypergraphs, namely the fact that in the former edges connect directly to nodes inside the equivalence class, whereas in the latter edges connect to the equivalence class itself.
Explicit discarding (weakening) nodes, depicted with black dots with a single wire inside the dashed boxes 
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.

The other steps, (c) to (e), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rules.

A typical way on encoding $\beta$-reduction in lambda-calculus in e-graphs is by explicit substitution~\cite{EggPaper,koehler2022sketchguided}: by adding a special substitution nodes and introduce rewrite rules for them.
After the reduction has been done these nodes serve no purpose and can be removed.
We show that $\beta$-reduction comes naturally as a composition when using string diagrams for enriched closed monoidal categories.
Consider an example in Figure~\ref{fig:e-graph-substitution} where successive iterations of an e-graph for a term $(\lambda x . (y + y) + x) 1$ are shown following the $\beta$-reduction.
First the subterm for an explicit substitution is introduced which is deemed equivalent to the top-level application term.
It is then rewritten using the rules for explicit substitution at the top of each column.
At the end all explicit substitution nodes can be removed.

Now consider an example from Figure~\ref{fig:e-string-substitution} which encodes the same $\beta$-reduction for the same term.
The $\lambda$-abstraction is represented by a rounded box and the bound variable is the dangling wire.
The application node is depicted as 
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=new edge style 2] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}. 
The reduction is just the re-attachment of wires.
Note how we avoided re-attachment for subterms as they did not use the bound variable wire, unlike in the case $[x \mapsto 1](y+y)$ before.

A common solution to replace explicit variable names in e-graphs is by using De Brujin indices~\cite{koehler2022sketchguided}.
The issue with this approach is that it interferes with sharing as equivalent subterms can be structurally different.
Consider the example $\lambda f . f ((\lambda x . f x) 2)$ which is represented as $\lambda . \%0 ((\lambda . \%1 \%0) 2)$ using De Brujin indices.
Notice how two different occurrences of the same bound variable $f$ are represented differently by $\%0$ and $\%1$ which means they can not be shared in an e-graph.
A string diagrammatic representation of the same term makes it possible for $f$ to be shared as demonstrated below.

\[
\adjustbox{scale=0.6}{
    \tikzfig{figures/de-brujin-string}
}
\]

>>>>

E-graphs can efficiently simultaneously represent several equivalent terms of an algebraic theory, generalising the DAG representation of terms to include equivalence classes.
The key concepts of e-graphs can be intuitively grasped from some simple (but non-trivial) examples, as shown in Figure~\ref{fig:e-graph-example}, a `Hello world!' of e-graphs~\cite{EggPaper}.

@@

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the approach in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}. 
Note that for the initial term, $(a*2)/2$ (subfigure (a)), which is already represented efficiently in the e-graph by sharing the node `2', the string diagram version shares this node explicitly, using a contraction node
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
			\node [style=none] (2) at (1, 4) {};
			\node [style=none] (3) at (2, 4) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
			\draw [bend right=45] (0) to (3.center);
			\draw [bend left=45] (0) to (2.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}
.

Nodes of an e-graph are encapsulated in dashed boxes indicating equivalence classes of nodes, or, more generally, of subgraphs rooted at these boxes.
A term graph is embedded as an e-graph by making each equivalence class have a single node.
The first rewrite rule, replacing multiplication by \emph{shift-left} creates the first non-unitary equivalence class, which includes the multiplication ($*$) and shift-left ($<\!\!<$) nodes,  now sharing the node `$a$'.
This second e-graph illustrates the other representational difference, besides sharing, between conventional e-graphs and e-hypergraphs.
In the former, edges connect directly to nodes inside the equivalence class, whereas in the latter edges connect to the equivalence class itself.
Explicit discarding (weakening) nodes, depicted with black dots with a single wire inside the dashed boxes 
\scalebox{0.4}{
	\begin{tikzpicture}
		\begin{pgfonlayer}{nodelayer}
			\node [style=vertex] (0) at (1.5, 3.5) {};
			\node [style=none] (1) at (1.5, 3) {};
		\end{pgfonlayer}
		\begin{pgfonlayer}{edgelayer}
			\draw (0) to (1.center);
		\end{pgfonlayer}
	\end{tikzpicture}
}, indicate which of the class-level edges are connected to which nodes inside the class.

The other steps, (c) to (e), represent further elaborations of the e-graph, or e-hypergraph, via the application of more rewrite rules.

A common way of dealing with $\beta$-reduction for the $\lambda$-calculus in e-graphs is by using special nodes for explicit substitution and associating them with particular rewrite rules~\cite{EggPaper,koehler2022sketchguided}.
The auxiliary nature of these nodes is revealed by the fact that after rewrites they must be removed from the graph. 
We will see how in the categorically-directed representation these nodes are not necessary because explicit substitution is simply modelled categorically as composition, a significant simplification.

\subsection{Conventional vs. categorical e-graphs with binding}

Before diving into the technical details let us compare how substitution works in conventional e-graphs as compared to our proposed approach.
In Figure~\ref{fig:e-graph-substitution} we show successive iterations of a conventional e-graph for $(\lambda x . (y + y) + x) 1$ following the $\beta$-reductions.
A subterm for an explicit substitution must be introduced, which is deemed equivalent to the top-level application term, to be rewritten using the rules for explicit substitution at the top of each column.
At the end, all explicit substitution nodes can be removed.

The same example is given in Figure~\ref{fig:e-string-substitution} using string diagrams. 
The $\lambda$-abstraction is represented by a rounded box and the bound variable is the dangling wire.
The application node is depicted as 
\adjustbox{scale=0.15}{\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=none] (0) at (1, 6.75) {};
		\node [style=none] (1) at (1.25, 7.75) {};
		\node [style=none] (2) at (2, 8) {};
		\node [style=none] (3) at (2.75, 7.75) {};
		\node [style=none] (4) at (3, 6.75) {};
		\node [style=none] (5) at (1.5, 6.75) {};
		\node [style=none] (6) at (2.5, 6.75) {};
		\node [style=none] (7) at (2, 9) {};
		\node [style=none] (8) at (1.5, 5.75) {};
		\node [style=none] (9) at (2.5, 5.75) {};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=new edge style 2] (3.center)
			 to [in=60, out=-30, looseness=0.75] (4.center)
			 to (0.center)
			 to [in=-150, out=120, looseness=0.75] (1.center)
			 to [in=-180, out=45, looseness=0.75] (2.center)
			 to [in=135, out=0, looseness=0.75] cycle;
		\draw (2.center) to (7.center);
		\draw (5.center) to (8.center);
		\draw (9.center) to (6.center);
	\end{pgfonlayer}
\end{tikzpicture}
}. 
The reduction is now simply the re-attachment of wires, and the overall process is visibly simpler. 
DRG UNCLEAR: Note how we avoided re-attachment for subterms as they did not use the bound variable wire, unlike in the case $[x \mapsto 1](y+y)$ before.

DRG UNCLEAR: A common solution to replace explicit variable names in e-graphs is by using De Brujin indices~\cite{koehler2022sketchguided}.
The issue with this approach is that it interferes with sharing as equivalent subterms can be structurally different.
Consider the example $\lambda f . f ((\lambda x . f x) 2)$ which is represented as $\lambda . \%0 ((\lambda . \%1 \%0) 2)$ using De Brujin indices.
Notice how two different occurrences of the same bound variable $f$ are represented differently by $\%0$ and $\%1$ which means they can not be shared in an e-graph.
A string diagrammatic representation of the same term makes it possible for $f$ to be shared as demonstrated below.

DRG USE WRAP-BOX. 
\[
\adjustbox{scale=0.6}{
    \tikzfig{figures/de-brujin-string}
}
\]
